---
title: "Clinical/Parasite Comparison for Vaccinees"
author: "Stephanie Wu"
date: "May 2, 2018"
output: word_document
---


```{r, echo=FALSE, warning=FALSE, message=FALSE}
source("C:/Users/Swu2/Documents/AADiversityIndices.R")
knitr::opts_chunk$set(fig.width=9, fig.height=12)
library(dplyr)
library(readr)
library(ape)
library(ade4)
library(seqinr)
library(RColorBrewer)
library(Biostrings)
library(reshape2)
library(ggplot2)
library(gridExtra)
library(ggpubr)

#vaccine 3D7 strain
vax3D7 <- "DENANANSAVKNNNNEEPSDKHIKEYLNKIQNSLSTEWSPCSVTCGNGIQVRIKPGSANKPKDELDYANDIEKKICKMEKCSSVFNVVNSSIGLI"

#load TEP data and clinical and parasite indicator files
setwd("T:/vaccine/rtss_malaria_sieve/")
TEP.data <- read_tsv("qdata/sequences/TEP.tsv")
marks.data.x <- read_tsv("adata/marks_data_x_sites.tsv")  #parasite positive cases
marks.data.c <- read_tsv("adata/marks_data_c_sites.tsv")  #clinical malaria cases

#subset for subjects age 5-17 months, separated into placebo and vaccine
clinical.data <- read_csv("adata/RTSSclinicalData.csv")
clinical.data <- subset(clinical.data, ageCateg=="[5-17] months")
clinical.placebo <- subset(clinical.data, vaccine==0)
clinical.vaccine <- subset(clinical.data, vaccine==1)

#TEP data for all participants age 5-17 months, as well as separated into vaccine and placebo groups
TEP.all <- TEP.data[TEP.data$subject %in% clinical.data$id,]
TEP.placebo <- TEP.data[TEP.data$subject %in% clinical.placebo$id,]
TEP.vaccine <- TEP.data[TEP.data$subject %in% clinical.vaccine$id,]

#number of distinct subjects in TEP data for participants age 5-17 months
subjects.TEP <- TEP.all %>% distinct(subject)
subjects.TEP <- subjects.TEP$subject

#number of distinct subjects in TEP placebo group for participants age 5-17 months
subjects.placebo <- TEP.placebo %>% distinct(subject)
subjects.placebo <- subjects.placebo$subject

#number of distinct subjects in TEP vaccine group for participants age 5-17 months
subjects.vaccine <- TEP.vaccine %>% distinct(subject)
subjects.vaccine <- subjects.vaccine$subject

#cases of clinical malaria for TEP placebo subjects aged 5-17 months
clinical <- TEP.data[TEP.data$subject %in% subjects.vaccine,]
clinical <- clinical[clinical$sample %in% marks.data.c$sample,]

#cases of parasite positivity for TEP placebo subjects aged 5-17 months
parasitepos <- TEP.data[TEP.data$subject %in% subjects.vaccine,]
parasitepos <- parasitepos[parasitepos$sample %in% marks.data.x$sample,]

#cases of both clinical malaria and parasite positivity for TEP placebo subjects aged 5-17 months
both <- clinical[clinical$subject %in% parasitepos$subject,]
both <- both[,1:11]

#number of distinct subjects with both clinical and parasite endpoints 
#for TEP placebo subjects aged 5-17 months
subjects.both <- both %>% distinct(subject)
subjects.both <- subjects.both$subject

# #number of distinct subjects
# n_distinct(clinical$subject)  #1181
# n_distinct(parasitepos$subject)  #286
# n_distinct(both$subject)   #161
```

### Comparison plots
**x = days between clinical malaria and parasite positive endpoint, y = parasite positive index value - clinical malaria index value. **


```{r, echo=FALSE, warning=FALSE, message=FALSE}
#load clinical/parasite index comparisons dataframe 
comparisons.df <- read.csv("C:/Users/Swu2/Documents/ClinicalParasiteCompVax", row.names = 1, header = TRUE)

#add columns for the differences between parasite pos at month 20 and clinical malaria
differences <- c("H0.diff", "H1.diff","H2.diff","CorSegSites.diff","GS.diff","Pi.diff","Mf.diff","Hamming.diff")
comparisons.df[differences] <- NA
for(i in 1:8){
  comparisons.df[,(i+16)] <- comparisons.df[,(i+8)] - comparisons.df[,i]
}

#add columns for the ratios between parasite pos at month 20 and clinical malaria
ratios <- c("H0.ratio", "H1.ratio","H2.ratio","CorSegSites.ratio","GS.ratio","Pi.ratio","Mf.ratio","Hamming.ratio")
comparisons.df[ratios] <- NA
for(i in 1:8){
  comparisons.df[,(i+24)] <- comparisons.df[,(i+8)]/comparisons.df[,i]
}

#start date: 14 days after 3rd vax
start.date <- clinical.data %>% filter(id %in% both$subject) %>% select(id, startdate)
colnames(start.date) <- c("subject","startdate")
#date of clinical malaria sample
clin.date <- clinical.data %>% filter(id %in% both$subject) %>% select(id, fdate)
colnames(clin.date) <- c("subject","clindate")
#date of month 20 visit (parasite positivity sample)
m20 <- clinical.data %>% filter(id %in% both$subject) %>% select(id, visitdateM20)
colnames(m20) <- c("subject","m20")

comparisons.df$subject <- as.integer(rownames(comparisons.df))
#add date columns to comparisons.df
comparisons.df <- left_join(comparisons.df, clin.date, by="subject")
comparisons.df <- left_join(comparisons.df, start.date, by="subject")
comparisons.df <- left_join(comparisons.df, m20, by="subject")
#date difference between parasite sample and clinical malaria column
comparisons.df$date.diff <- as.numeric(comparisons.df$m20 - comparisons.df$clindate)
#date difference between clinical malaria and start date column
comparisons.df$date.diff.clin <- as.numeric(comparisons.df$clindate - comparisons.df$startdate)


# #write comparison plot graphics to pdf 
# pdf("C:/Users/Swu2/Documents/parasite.pdf")
p1 <- ggplot(comparisons.df, aes(x=date.diff, y=H0.diff)) + geom_point() + stat_smooth(method = loess)
p2 <- ggplot(comparisons.df, aes(x=date.diff, y=H1.diff)) + geom_point() + stat_smooth(method = loess)
p3 <- ggplot(comparisons.df, aes(x=date.diff, y=H2.diff)) + geom_point() + stat_smooth(method = loess)
p4 <- ggplot(comparisons.df, aes(x=date.diff, y=CorSegSites.diff)) + geom_point() + stat_smooth(method = loess)
p5 <- ggplot(comparisons.df, aes(x=date.diff, y=GS.diff)) + geom_point() + stat_smooth(method = loess)
p6 <- ggplot(comparisons.df, aes(x=date.diff, y=Pi.diff)) + geom_point() + stat_smooth(method = loess)
p7 <- ggplot(comparisons.df, aes(x=date.diff, y=Mf.diff)) + geom_point() + stat_smooth(method = loess)
p8 <- ggplot(comparisons.df, aes(x=date.diff, y=Hamming.diff)) + geom_point() + stat_smooth(method = loess)
grid.arrange(p1,p2,p3,p4, nrow=2)
grid.arrange(p5,p6,p7,p8, nrow=2)
# dev.off()

# for(i in 1:8){
# #print(t.test(comparisons.df[,i],comparisons.df[,i+8], alternative="two.sided", paired=TRUE))
# print(summary(comparisons.df[,i+16]))
# }
```

###Comparing changes in number of haplotypes and number of reads between clinical malaria samples and parasite positivity samples 

The number of haplotypes and the number of reads are both higher on average for parasite positivity samples compared to clinical malaria samples. Differences in number of haplotypes may be in part due to differences in number of reads, despite our fringe-trimming efforts to correct sample size bias. 

**Clinical malaria cases (top row is number of haplotypes, bottom row is number of reads)**
```{r, echo=FALSE, message=FALSE, warning=FALSE}
#haplotype and reads summary
#clinical cases
num.haplotypes <- numeric(length(subjects.both))
num.reads <- numeric(length(subjects.both))
for(i in 1:length(subjects.both)){
  sub <- subjects.both[i]
  data <- clinical %>% filter(subject==sub) %>% select(pep_sequence, reads)
  count <- as.vector(data$reads)
  #fringe-trimming
  cutoff <- qbinom(0.9, size=sum(count), prob = 0.01, lower.tail=TRUE)
  for(j in 1:length(count)){
    if(count[j] < cutoff){
      count[j] <- NA
    }
  }
  count <- na.omit(count)
  num.haplotypes[i] <- length(count)
  num.reads[i] <- sum(count)
}
summary(num.haplotypes)
summary(num.reads)
```

**Parasite positivity cases**
```{r, echo=FALSE, message=FALSE, warning=FALSE}
#parasite positive cases
num.haplotypes <- numeric(length(subjects.both))
num.reads <- numeric(length(subjects.both))
for(i in 1:length(subjects.both)){
  sub <- subjects.both[i]
  data <- parasitepos %>% filter(subject==sub) %>% select(pep_sequence, reads)
  count <- as.vector(data$reads)
  #fringe-trimming
  cutoff <- qbinom(0.9, size=sum(count), prob = 0.01, lower.tail=TRUE)
  for(j in 1:length(count)){
    if(count[j] < cutoff){
      count[j] <- NA
    }
  }
  count <- na.omit(count)
  num.haplotypes[i] <- length(count)
  num.reads[i] <- sum(count)
}
summary(num.haplotypes)
summary(num.reads)
```

###Summary statistics comparison between clinical and parasite samples
**"mean.c" refers to the mean value over clinical samples. "mean.x" refers to mean value over parasite samples. Compare adjacent rows to compare clinical with parasite.**


Although these results do not tell us about time-dependent change in these indices, they do suggest the indices have changed between the two samples. In general, the parasite positive samples have higher numbers for the indices, which suggests more heterogeneity in the samples. 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
##summary stats comparison
mean.c <- apply(comparisons.df[1:8], 2, function(x) mean(x))
mean.x <- apply(comparisons.df[9:16], 2, function(x) mean(x))
min.c <- apply(comparisons.df[1:8], 2, function(x) min(x))
min.x <- apply(comparisons.df[9:16], 2, function(x) min(x))
max.c <- apply(comparisons.df[1:8], 2, function(x) max(x))
max.x <- apply(comparisons.df[9:16], 2, function(x) max(x))
median.c <- apply(comparisons.df[1:8], 2, function(x) median(x))
median.x <- apply(comparisons.df[9:16], 2, function(x) median(x))
sd.c <- apply(comparisons.df[1:8], 2, function(x) sd(x))
sd.x <- apply(comparisons.df[9:16], 2, function(x) sd(x))
IQR.c <- apply(comparisons.df[1:8], 2, function(x) IQR(x))
IQR.x <- apply(comparisons.df[9:16], 2, function(x) IQR(x))

df <- data.frame(mean.c, mean.x, min.c, min.x, max.c, max.x, median.c, median.x, sd.c, sd.x, IQR.c, IQR.x)
rownames(df) <- c("H(0)","H(1)","H(2)","CorSegSites","GS","Rao","Mf","Hamming")

t(df)

pp <- t(df[,c(2,4,6,8,10,12)])
clin <- t(df[, c(1,3,5,7,9,11)])
new <- as.data.frame(rbind(clin, pp))
new$stat <- rep(c("mean","min","max","median","sd","IQR"),times=2)
new$label <- c(rep("clin", times=6), rep("pp", times=6))
new <- melt(new)
ggplot(new, aes(y=value, x=stat, color=label, fill=label)) + 
  geom_bar(stat="identity", position=position_dodge()) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  facet_wrap(~variable, scale="free")
```

###Histogram and boxplot comparison between clinical samples and parasite samples

These graphs also suggest the parasite positivity samples have more haplotypes, more equal distribution of haplotype frequencies, and more pairwise mutations. These show increased heterogeneity in the parasite positivity samples.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
###Histogram and boxplot comparison
clin <- comparisons.df[,1:8]
clin$label <- "clinical"
pp <- comparisons.df[,9:16]
pp$label <- "parasite"
colnames(pp) <- colnames(clin)
combined <- rbind(clin, pp)
colnames(combined) <- c("H0","H1","H2","CorSegSites","GS","Rao","Mf","Hamming","label")

p1 <- ggplot(combined, aes(x=H0, y=..count../sum(..count..), color=label, fill=label)) + geom_histogram(position="identity", alpha=0.5) + ylab("Density")
p2 <- ggplot(combined, aes(x=H1, y=..count../sum(..count..),color=label, fill=label)) + geom_histogram(position="identity", alpha=0.5) + ylab("Density")
p3 <- ggplot(combined, aes(x=H2, y=..count../sum(..count..),color=label, fill=label)) + geom_histogram(position="identity", alpha=0.5) + ylab("Density")
p4 <- ggplot(combined, aes(x=CorSegSites, y=..count../sum(..count..),color=label, fill=label)) + geom_histogram(position="identity", alpha=0.5) + ylab("Density")
p5 <- ggplot(combined, aes(x=GS, y=..count../sum(..count..),color=label, fill=label)) + geom_histogram(position="identity", alpha=0.5) + ylab("Density")
p6 <- ggplot(combined, aes(x=Rao, y=..count../sum(..count..),color=label, fill=label)) + geom_histogram(position="identity", alpha=0.5) + ylab("Density")
p7 <- ggplot(combined, aes(x=Mf, y=..count../sum(..count..),color=label, fill=label)) + geom_histogram(position="identity", alpha=0.5) + ylab("Density")
p8 <- ggplot(combined, aes(x=Hamming, y=..count../sum(..count..),color=label, fill=label)) + geom_histogram(position="identity", alpha=0.5) + ylab("Density")

ggarrange(p1, p2, p3, p4, ncol=2, nrow=2, common.legend = TRUE, legend="top")
ggarrange(p5, p6, p7, p8, ncol=2, nrow=2, common.legend = TRUE, legend="top")

#boxplots
boxplotdata <- melt(combined)
ggplot(boxplotdata, aes(x=variable, y=value, fill=label)) + geom_boxplot() + facet_wrap(~variable, scale="free")

```
